Profit Optimisation in Heterogeneous 2-link Parallel Games

### 2024-03-18 thesis/article.tex
Thesis: Refactors split properties to different lemmas

\begin{lemma}
	Given a 2-link parallel heterogeneous game with distribution function $\alpha$ and time-money sensitivity split function $\as$, the following hold:
	\begin{enumerate}[(i)]
		\item $\as(t) > 0$
		\item if $x_1(t), x_2(t) > 0$ then $\as(t) \in [\alpha(0), \alpha(1)]$
		\item if $0 < x_1(t) < x_2(t)$ then $\alpha(x_1(t)) \le \as(t) \le \alpha(x_1(t) + \epsilon), \forall \epsilon > 0$ arbitrarily small
		\item if $t, t^\prime$ tolls such that $\frac{t_1 - t_2}{t_1^\prime - t_2^\prime} > 0$ then
			\begin{itemize}
				\item $\as(t) \ge \as(t^\prime)$ iff $\frac{t_1 - t_2}{t_1^\prime - t_2^\prime} \le 1$
				\item $\as(t) \le \as(t^\prime)$ iff $\frac{t_1 - t_2}{t_1^\prime - t_2^\prime} \ge 1$
			\end{itemize}
		\item $\as(t) \rightarrow 0$ iff $|t_1 - t_2| \rightarrow \infty$
		\item $\lim_{t_1 \rightarrow t_2^-} \as(t) = \alpha(x_1), \lim_{t_1 \rightarrow t_2^+} \as(t) = \alpha(x_2)$, where ($x_1, x_2)$ the optimal flow under no tolls
	\end{enumerate}
\end{lemma}

\begin{proof}
	$ $ % Just to make a line break after "Proof".
	\begin{enumerate}[(i)]
		\item Obviously $\as(t) \ne 0$ since $\as(t) = 0 \Rightarrow l_1(x_1(t)) = l_2(x_2(t))$ which is impossible in a heterogeneous setup.
		Then we only need to notice that $l_1(x_1(t)) < l_2(x_2(t)) \iff t_1 > t_2$ and we get $\as(t) > 0$ from definition.
		\item Since $x_1(t), x_2(t) > 0$ then $\exists p_1, p_2 \in [0, 1]$ each on a different edge, for which we assume w.l.o.g. that $\alpha(p_1) \le \alpha(p_2)$.
		It holds that $\alpha(p_1) \le \as(t) \le \alpha(p_2)$ so with $\alpha$ being non-decreasing it follows
		\[\alpha(0) \le \alpha(p_1) \le \as(t) \le \alpha(p_2) \le \alpha(1) \Rightarrow \as(t) \in [\alpha(0), \alpha(1)]\]
		\item We know that $x_1(t) < x_2(t)$ means that edge 1 has the lower split.
		Therefore we have
		\[\alpha(p) \le \as(t), \forall p \in [0, x_1(t)] \Rightarrow \alpha(x_1(t)) \le \as(t)\]
		Likewise for edge 2 we have
		\[\as(t) \le \alpha(p), \forall p \in (x_1(t), 1] \Rightarrow \as(t) \le \alpha(x_1(t) + \epsilon), \forall \epsilon \in (0, 1 - x_1(t))\]
		\item The condition
		\begin{equation}
			\frac{t_1 - t_2}{t_1^\prime - t_2^\prime} > 0
			\label{eq:split_property_cond_tolls_0}
		\end{equation}
		ensures that the two tolls provide splits with same edges in their respective lower and higher sections.
		We show this for $t_1 > t_2$ and likewise it also holds for $t_1 < t_2$.
		\[t_1 > t_2 \Leftrightarrow t_1 - t_2 > 0 \LeftrightarrowArg{(\ref{eq:split_property_cond_tolls_0})} t_1^\prime - t_2^\prime > 0 \Leftrightarrow t_1^\prime > t_2^\prime\]
		%Also note that this property extends to the latencies, giving us
		%\begin{equation}
		%	\frac{l_1(x_1(t)) - l_2(x_2(t))}{l_1(x_1^\prime(t)) - l_2(x_2^\prime(t))} > 0
		%	\label{eq:split_property_cond_latencies_0}
		%\end{equation}
		Then moving on to the first case, we first see that $\frac{t_1 - t_2}{t_1^\prime - t_2^\prime} \le 1 \Leftrightarrow |t_1 - t_2| \le |t_1^\prime - t_2^\prime|$ (opposite direction is ensured by $(\ref{eq:split_property_cond_tolls_0})$), i.e. the toll difference is increasing.
		Assume w.l.o.g. that $x_1 < x_2$.

@@

		If $x_1(t) > 0$ then we know [TODO: do we?] that the latency difference might also increase, with flow from the lower latency edge, if any, moving towards the slower one, i.e. $x_1(t) \ge x_1(t^\prime)$.
		If no flow moves then we can easily show that $\as(t) \ge \as(t^\prime)$ similarly to the case with $x_1(t) = 0$.
		Otherwise $x_1(t) > x_1(t^\prime) \Rightarrow \exists \epsilon > 0: x_1(t) = x_1(t^\prime) + \epsilon$.
		%\[|t_1 - t_2| \le |t_1^\prime - t_2^\prime| \Leftrightarrow |l_1(x_1(t)) - l_2(x_2(t))| \le |l_1(x_1(t^\prime)) - l_2(x_2(t^\prime))|\]
		%\begin{equation}
		%	|l_1(x_1(t)) - l_2(x_2(t))| \le |l_1(x_1(t^\prime)) - l_2(x_2(t^\prime))| \LeftrightarrowArg{(\ref{eq:split_property_cond_latencies_0})} \frac{l_1(x_1(t)) - l_2(x_2(t))}{|l_1(x_1(t^\prime)) - l_2(x_2(t^\prime))|} \le 1
		%	\label{eq:split_property_cond_latencies_1}
		%\end{equation}
		Finally, by carefully applying property (ii) we get
		\[\as(t^\prime) \le \alpha(x_1(t^\prime) + \epsilon) = \alpha(x_1(t)) \le \as(t)\]

		All steps can be followed backwards as well, therefore the opposite also holds.

		For the second case, we see that the toll difference decreases, and following respectively the same arguments we can show that $\as(t) \le \as(t^\prime)$.
		\item If $\alpha(0) > 0$ then...
	\end{enumerate}
\end{proof}

>>>>

We start with some basic properties.
\begin{lemma}
	\label{lemma:split_basic}
	Given a 2-link parallel heterogeneous game with distribution function $\alpha$ and time-money sensitivity split function $\as$, the following hold:
	\begin{enumerate}[(i)]
		\item $\as(t) > 0$
		\item if $x_1(t), x_2(t) > 0$ then $\as(t) \in [\alpha(0), \alpha(1)]$
		\item if $0 < x_1(t) < x_2(t)$ then $\alpha(x_1(t)) \le \as(t) \le \alpha(x_1(t) + \epsilon), \forall \epsilon > 0$ arbitrarily small
	\end{enumerate}
\end{lemma}

\begin{proof}
	$ $ % Just to make a line break after "Proof".
	\begin{enumerate}[(i)]
		\item Obviously $\as(t) \ne 0$ since $\as(t) = 0 \Rightarrow l_1(x_1(t)) = l_2(x_2(t))$ which is impossible in a heterogeneous setup.
		Then we only need to notice that $l_1(x_1(t)) < l_2(x_2(t)) \iff t_1 > t_2$ and we get $\as(t) > 0$ from definition.
		\item Since $x_1(t), x_2(t) > 0$ then $\exists p_1, p_2 \in [0, 1]$ each on a different edge, for which we assume w.l.o.g. that $\alpha(p_1) \le \alpha(p_2)$.
		It holds that $\alpha(p_1) \le \as(t) \le \alpha(p_2)$ so with $\alpha$ being non-decreasing it follows
		\[\alpha(0) \le \alpha(p_1) \le \as(t) \le \alpha(p_2) \le \alpha(1) \Rightarrow \as(t) \in [\alpha(0), \alpha(1)]\]
		\item We know that $x_1(t) < x_2(t)$ means that edge 1 has the lower split.
		Therefore we have
		\[\alpha(p) \le \as(t), \forall p \in [0, x_1(t)] \Rightarrow \alpha(x_1(t)) \le \as(t)\]
		Likewise for edge 2 we have
		\[\as(t) \le \alpha(p), \forall p \in (x_1(t), 1] \Rightarrow \as(t) \le \alpha(x_1(t) + \epsilon), \forall \epsilon \in (0, 1 - x_1(t))\]
	\end{enumerate}
\end{proof}

The next lemma will help us describe how $\as(t)$ is changing in relation to toll difference (and in extend to latency difference).
\begin{lemma}
	Given a 2-link parallel heterogeneous game with time-money sensitivity split function $\as$ and tolls $t, t^\prime$ such that $\frac{t_1 - t_2}{t_1^\prime - t_2^\prime} > 0$, the following hold:
	\begin{enumerate}[(i)]
		\item $\as(t) \ge \as(t^\prime)$ iff $\frac{t_1 - t_2}{t_1^\prime - t_2^\prime} \le 1$
		\item $\as(t) \le \as(t^\prime)$ iff $\frac{t_1 - t_2}{t_1^\prime - t_2^\prime} \ge 1$
	\end{enumerate}
\end{lemma}

\begin{proof}
	Before we get into the propositions, notice that the condition
	\begin{equation}
		\frac{t_1 - t_2}{t_1^\prime - t_2^\prime} > 0
		\label{eq:split_property_cond_tolls_0}
	\end{equation}
	ensures that the two tolls provide splits with same edges in their respective lower and higher sections.
	We show this for $t_1 > t_2$ and likewise it also holds for $t_1 < t_2$.
	\[t_1 > t_2 \Leftrightarrow t_1 - t_2 > 0 \LeftrightarrowArg{(\ref{eq:split_property_cond_tolls_0})} t_1^\prime - t_2^\prime > 0 \Leftrightarrow t_1^\prime > t_2^\prime\]
	%Also note that this property extends to the latencies, giving us
	%\begin{equation}
	%	\frac{l_1(x_1(t)) - l_2(x_2(t))}{l_1(x_1^\prime(t)) - l_2(x_2^\prime(t))} > 0
	%	\label{eq:split_property_cond_latencies_0}
	%\end{equation}
	Now moving on to the prepositions, we focus on proving $(i)$ in the right-to-left direction; then the same steps can be followed backwards to prove left-to-right.
	Finally proving $(ii)$ will be much easier as there are many steps symmetric to the ones for $(i)$.

	\paragraph{$(i)$}
	Begin by noticing that $\frac{t_1 - t_2}{t_1^\prime - t_2^\prime} \le 1 \Leftrightarrow |t_1 - t_2| \le |t_1^\prime - t_2^\prime|$ (opposite direction is ensured by $(\ref{eq:split_property_cond_tolls_0})$), i.e. the toll difference is increasing.
	Assume w.l.o.g. that $x_1 < x_2$.

@@

	If $x_1(t) > 0$ then we know [TODO: do we?] that the latency difference might also increase, with flow from the lower latency edge, if any, moving towards the slower one, i.e. $x_1(t) \ge x_1(t^\prime)$.
	%\[|t_1 - t_2| \le |t_1^\prime - t_2^\prime| \Leftrightarrow |l_1(x_1(t)) - l_2(x_2(t))| \le |l_1(x_1(t^\prime)) - l_2(x_2(t^\prime))|\]
	%\begin{equation}
	%	|l_1(x_1(t)) - l_2(x_2(t))| \le |l_1(x_1(t^\prime)) - l_2(x_2(t^\prime))| \LeftrightarrowArg{(\ref{eq:split_property_cond_latencies_0})} \frac{l_1(x_1(t)) - l_2(x_2(t))}{|l_1(x_1(t^\prime)) - l_2(x_2(t^\prime))|} \le 1
	%	\label{eq:split_property_cond_latencies_1}
	%\end{equation}
	If no flow moves then we can easily show that $\as(t) \ge \as(t^\prime)$ similarly to the case with $x_1(t) = 0$.
	Otherwise we have $x_1(t) > x_1(t^\prime) \Rightarrow \exists \epsilon > 0: x_1(t) = x_1(t^\prime) + \epsilon$ and by carefully applying property $\ref{lemma:split_basic}(ii)$ we get
	\[\as(t^\prime) \le \alpha(x_1(t^\prime) + \epsilon) = \alpha(x_1(t)) \le \as(t)\]

	\paragraph{$(ii)$}
	Symmetrically to $(i)$ in this case the toll difference is decreasing, and therefore flow from the higher latency edge will move towards the lower one, i.e. $x_1(t) \le x_1(t^\prime)$ (assuming again $x_1 < x_2$).
	With a similar argument we then get $\as(t) \le \as(t^\prime)$.
\end{proof}

The next lemma gives us some insight of the behavior of $\as(t)$ around its edge values.
\begin{lemma}
	Given a 2-link parallel heterogeneous game with distribution function $\alpha$ and time-money sensitivity split function $\as$, the following hold:
	\begin{enumerate}[(i)]
		\item $\as(t) \rightarrow 0$ iff $|t_1 - t_2| \rightarrow \infty$
		\item $\lim_{t_1 \rightarrow t_2^-} \as(t) = \alpha(x_1), \lim_{t_1 \rightarrow t_2^+} \as(t) = \alpha(x_2)$, where ($x_1, x_2)$ the optimal flow under no tolls
	\end{enumerate}
\end{lemma}

\begin{proof}
	\begin{enumerate}[(i)]
		\item If $\alpha(0) > 0$ then...
		\item 
	\end{enumerate}
\end{proof}
